The past and future of visualization for computing in science and engineering

### 2020-03-27 CombaWeiskopf2020/paper.tex
edits section 1

\subsubsection{Multidimensional and Multivariate Data Visualization}
% 4 papers:
% Paulovich, F., Silva, C., & Nonato, L. (2012). User-centered multidimensional projection techniques. Computing in Science and Engineering, 14(4), 74-81.
% Da Silva, R., Rauber, P., & Telea, A. (1. 9 2016). Beyond the third dimension: Visualizing high-dimensional data with projections. Computing in Science and Engineering, 18(5), 98-107.
% Doleisch, H., & Hauser, H. (3 2012). Interactive visual exploration and analysis of multivariate simulation data. Computing in Science and Engineering, 14(2), 70-76.
% Heinrich, J., & Weiskopf, D. (2015). Parallel coordinates for multidimensional data visualization: Basic concepts. Computing in Science and Engineering, 17(3), 70-76.

Multidimensional and multivariate data is almost omnipresent in computational science and engineering. Accordingly, we had several papers addressing this topic.
The paper by Helmut Doleisch and Helwig Hauser~\cite{Doleisch2012} nicely shows how the combination of 2D and 3D visualization, flexible degree-of-interest definition, and interactive exploration leads to an effective visual analysis of complex simulation data. They discuss, for example, how multiple simulation attributes from computational fluid dynamics can be visually analyzed. Figure~\ref{fig:simvis} shows a screenshot from their software system, illustrating how different visualization views are integrated in order to support a user-oriented exploration and analysis process.

\begin{figure*}
    \begin{center}    
        \includegraphics[width=0.9\textwidth]{simvis.png}
        \caption{Example of multivariate data visualization with the SimVis system~\cite{Doleisch2012}. This screenshot shows the windflow around a building, where turbulent parts are emphasized using a vortex detection method. Image: \copyright{} 2012 IEEE, reprinted from Doleisch and Hauser~\cite{Doleisch2012}.
  \label{fig:simvis}}        
    \end{center}
\end{figure*}

Other papers for this theme covered multidimensional projection techniques that reduce multidimensional data to 2D data for visualization purposes. One paper specifically included user-centered control into the projection techniques. Another paper followed a different approach, employing parallel coordinates to show multiple data dimensions simultaneously. Overall, the papers in this category come with quite generic visualizations that can be useful in many areas of science and engineering.

\subsubsection{Flow Visualization}
% 3+1 papers:
%Netzel, R., & Weiskopf, D. (2013). Texture-based flow visualization. Computing in Science and Engineering, 15(6), 96-102.
%Karch, G., Sadlo, F., Weiskopf, D., Hansen, C., Li, G., & Ertl, T. (2012). Dye-based flow visualization. Computing in Science and Engineering, 14(6), 80-86.
%Weinkauf, T., & Theisel, H. (2012). Flow visualization and analysis using streak and time lines. Computing in Science and Engineering, 14(5), 78-84.
% also in HCI: Englund, R., Palmerius, K., Hotz, I., & Ynnerman, A. (1. 5 2018). Touching Data: Enhancing Visual Exploration of Flow Data with Haptics. Computing in Science and Engineering, 20(3), 89-100.

This topic area is a classic one in scientific visualization, considering input from flow simulation and experiments alike. We pick one of the VizCorner papers as an example: Grzegorz Karch and colleagues~\cite{Karch2012} described methods for high-quality computational flow visualization utilizing the metaphor of dye advection. Experimental flow visualization has a long tradition of dye-based observations. With its computational counterpart, we can use the same idea now in a computer-based visualization, see Figure~\ref{fig:karman} as a typical example.

\begin{figure*}
	\begin{center}	
		\includegraphics[width=0.9\textwidth]{karman_weno3.png}
		\caption{Computational dye advection~\cite{Karch2012}, applied to a 2D simulation of a von K{\'a}rm{\'a}n vortex street. Image: \copyright{} 2012 IEEE, reprinted from Karch et al.~\cite{Karch2012}.
  \label{fig:karman}}		
	\end{center}
\end{figure*}

Other papers described alternative visual representations of flow, for example, by using textures or streak and timelines. Another paper highlighted interaction techniques with haptics that let users virtually touch the data.

\subsubsection{Human-Computer Interaction and Storytelling}

@@

The latter paper leads us to the next theme that focuses on the human-centered aspects of visualization. In general, interaction plays an essential role in visualization; for example, in order to support scientists in exploring their simulation data. Other VizCorner articles highlighted new developments in eye tracking for interactive visualization or the role of human visual perception. Finally, Franz Sauer and colleagues~\cite{Sauer2016} went all the way to discussing visualization as a means of storytelling that can address a wide range of audiences, from domain scientists to general scientists to the public. For the latter one, they showcased an example of an interactive museum exhibit that visualized plankton populations to museum visitors, see Figure~\ref{fig:museum}.

@@

\subsubsection{Astronomy and Astrophysics}

@@

The above example of the museum exhibit links to other visualization activities for public outreach: 
Miguel Aragon-Calvo and Mark SubbaRao~\cite{Aragon-Calvo2015} described a virtual flight through the universe that addresses and balances scientific and artistic goals. For example, they discussed how they adapted the virtual size of galaxies in order to address the issue of vast space between those in the visualization. Figure~\ref{fig:galaxies} shows a snapshot of a flight through the universe.

@@

Another VizCorner paper looked into the issue of reconstructing 3D models of astronomical nebulae from 2D images along with the efficient volume rendering of such nebulae. Other papers discussed approaches to visualizing special relativity and general relativity.

\subsubsection{Further Applications and Visualization Techniques}
%Salinet, J., Oliveira, G., Vanheusden, F., Comba, J., Ng, G., & Schlindwein, F. (2013). Visualizing intracardiac atrial fibrillation electrograms using spectral analysis. Computing in Science and Engineering, 15(2), 79-87.
%Santos, E., Poco, J., Wei, Y., Liu, S., Cook, B., Williams, D., & Silva, C. (2013). UV-CDAT: Analyzing climate datasets from a user's perspective. Computing in Science and Engineering, 15(1), 94-103.
%Michalski, M., Rieth, M., Kempf, A., & Krüger, J. (1. 11 2017). CoFlaVis: A visualization system for pulverized coal flames. Computing in Science and Engineering, 19(6), 72-78.

The field of astronomy and astrophysics is only one relevant application area for visualization. The VizCorner acknowledges the breadth of possible applications. Other application examples covered in previous papers include the visualization of climate data sets, simulation of coal combustion processes in the form of pulverized coal flames, or the visualization of intracardiac atrial fibrillation electrograms. These examples show the full range of visualization applications from science to engineering to medical science. Also, they showcase the practical relevance of visualization---one of the goals of the VizCorner.

With multidimensional and multivariate visualization, flow visualization, and human-computer interaction, we have already touched some relevant visualization themes that build a versatile foundation for many applications. However, the VizCorner is not restricted to just these themes. Previous papers covered approaches to direct analytical support in visualization, advanced volume rendering, efficient parallel processing, mathematical modeling, or incorporating text in visualization.  %---showing the breadth in visualization methods relevant for science and engineering.

@@

\subsubsection{Summary and Reflections}

This very brief tour through past VizCorner articles shows the breadth of basic visualization techniques and the full range of exciting applications alike. We have seen some themes covered in more depth, such as flow visualization or multivariate data visualization---these are directly driven by typical visualization needs in science and engineering. However, the VizCorner also tries to bring in trends from the visualization research community that might have a rather long-term impact on science and engineering. For example, advanced human-computer interaction techniques might eventually improve how engineers or scientists work with their visual-interactive tools.

Two VizCorner articles~\cite{Scheidegger2011,Comba2017} have already provided an overview of trends in visualization and their impact on computing in science and engineering. They reported on the 2011 and 2016 IEEE VIS Conferences (wwww.ieeevis.org), which is the premier venue for visualization research. The two papers also included research trends that were present at that time. In particular, the 2016 IEEE VIS report~\cite{Comba2017} identified some trends that, in a similar form, are still relevant. Let us have a closer look at current trends in visualization research.

\section{Trends in Visualization}

>>>>

\subsection*{Multidimensional and multivariate data visualization}
% 4 papers:
% Paulovich, F., Silva, C., & Nonato, L. (2012). User-centered multidimensional projection techniques. Computing in Science and Engineering, 14(4), 74-81.
% Da Silva, R., Rauber, P., & Telea, A. (1. 9 2016). Beyond the third dimension: Visualizing high-dimensional data with projections. Computing in Science and Engineering, 18(5), 98-107.
% Doleisch, H., & Hauser, H. (3 2012). Interactive visual exploration and analysis of multivariate simulation data. Computing in Science and Engineering, 14(2), 70-76.
% Heinrich, J., & Weiskopf, D. (2015). Parallel coordinates for multidimensional data visualization: Basic concepts. Computing in Science and Engineering, 17(3), 70-76.

Multidimensional and multivariate data is almost omnipresent in computational science and engineering. 
Helmut Doleisch and Helwig Hauser~\cite{Doleisch2012} nicely showed how the combination of 2D and 3D visualization, flexible degree-of-interest definition, and interactive exploration lead to effective visual analysis of complex simulation data. They discussed how multiple simulation attributes from computational fluid dynamics can be visually analyzed. Figure~\ref{fig:simvis} shows a screenshot from their software system, illustrating how different visualization views are integrated in order to support a user-oriented exploration and analysis process.

\begin{figure*}
    \begin{center}    
        \includegraphics[width=0.9\textwidth]{simvis.png}
        \caption{Example of multivariate data visualization with the SimVis system. This screenshot shows the windflow around a building, where turbulent parts are emphasized using a vortex detection method. Image: \copyright{} 2012 IEEE, reprinted from Doleisch and Hauser~\cite{Doleisch2012}.
  \label{fig:simvis}}        
    \end{center}
\end{figure*}

The VizCorner has also covered projection techniques that reduce multidimensional data to 2D data for visualization purposes, including user-centered control of the projection techniques, and showing multiple data dimensions simultaneously via parallel coordinates. Papers in this category often include generic visualizations that can be useful in many areas of science and engineering.

\subsubsection*{Flow visualization}
% 3+1 papers:
%Netzel, R., & Weiskopf, D. (2013). Texture-based flow visualization. Computing in Science and Engineering, 15(6), 96-102.
%Karch, G., Sadlo, F., Weiskopf, D., Hansen, C., Li, G., & Ertl, T. (2012). Dye-based flow visualization. Computing in Science and Engineering, 14(6), 80-86.
%Weinkauf, T., & Theisel, H. (2012). Flow visualization and analysis using streak and time lines. Computing in Science and Engineering, 14(5), 78-84.
% also in HCI: Englund, R., Palmerius, K., Hotz, I., & Ynnerman, A. (1. 5 2018). Touching Data: Enhancing Visual Exploration of Flow Data with Haptics. Computing in Science and Engineering, 20(3), 89-100.

A classic research area, flow visualization considers input from flow simulation and experiments alike. Our pick from the VizCorner is by Grzegorz Karch and colleagues~\cite{Karch2012}, who described methods for high-quality flow visualization utilizing the metaphor of dye advection. Experimental flow visualization has a long tradition of dye-based observations. Its computational counterpart uses the same idea now in a computer-based visualization; see an example in Figure~\ref{fig:karman}.

\begin{figure*}
	\begin{center}	
		\includegraphics[width=0.9\textwidth]{karman_weno3.png}
		\caption{Computational dye advection applied to a 2D simulation of a von K{\'a}rm{\'a}n vortex street. Image: \copyright{} 2012 IEEE, reprinted from Karch et al.~\cite{Karch2012}.
  \label{fig:karman}}		
	\end{center}
\end{figure*}

The VizCorner has covered alternative visual representations of flow using textures or streak and timelines, 
and interaction techniques with haptics that let users virtually touch the data.

\subsubsection*{Human-computer interaction and storytelling}

@@

Another theme in the VizCorner focuses on the human-centered aspects of visualization. Interaction plays an essential role in visualization, supporting scientists in exploring their simulation data. The VizCorner has highlighted new developments in eye tracking for interactive visualization or the role of human visual perception. Franz Sauer and colleagues~\cite{Sauer2016} went all the way to discussing visualization as a means of storytelling that can address a wide range of audiences, from domain scientists to general scientists to the public. They showcased an example of an interactive museum exhibit that visualized plankton populations for visitors; see Figure~\ref{fig:museum}.

@@

\subsubsection{Astronomy and astrophysics}

@@

The museum exhibit mentioned above links to other visualization activities for public outreach: 
Miguel Aragon-Calvo and Mark SubbaRao~\cite{Aragon-Calvo2015} described a virtual flight through the universe that addresses and balances scientific and artistic goals. They discussed how they adapted the virtual size of galaxies in order to address the issue of vast space between them in the visualization. Figure~\ref{fig:galaxies} shows a snapshot of a flight through the universe.

@@

The VizCorner also looked into the issue of reconstructing 3D models of astronomical nebulae from 2D images along with the efficient volume rendering of such nebulae. Other papers discussed approaches to visualizing special relativity and general relativity.

\subsubsection{Other applications and visualization techniques}
%Salinet, J., Oliveira, G., Vanheusden, F., Comba, J., Ng, G., & Schlindwein, F. (2013). Visualizing intracardiac atrial fibrillation electrograms using spectral analysis. Computing in Science and Engineering, 15(2), 79-87.
%Santos, E., Poco, J., Wei, Y., Liu, S., Cook, B., Williams, D., & Silva, C. (2013). UV-CDAT: Analyzing climate datasets from a user's perspective. Computing in Science and Engineering, 15(1), 94-103.
%Michalski, M., Rieth, M., Kempf, A., & Krüger, J. (1. 11 2017). CoFlaVis: A visualization system for pulverized coal flames. Computing in Science and Engineering, 19(6), 72-78.

The VizCorner encompasses the breadth of possible applications of visualization. We have covered the visualization of climate data sets, simulation of coal combustion processes in the form of pulverized coal flames, or the visualization of intracardiac atrial fibrillation electrograms. Showing the full range of visualization applications from science to engineering to medical science, and the practical relevance of visualization, is among the goals of the VizCorner.

With multidimensional and multivariate visualization, flow visualization, and human-computer interaction, we have already touched some relevant visualization themes that build a versatile foundation for many applications. The VizCorner is not restricted to these themes, and we have covered approaches to direct analytical support in visualization, advanced volume rendering, efficient parallel processing, mathematical modeling, or incorporating text in visualization.  %---showing the breadth in visualization methods relevant for science and engineering.

@@

\subsubsection{Summary and reflections}

This brief tour through past VizCorner articles shows the breadth of basic visualization techniques and the full range of exciting applications. We have seen some themes covered in more depth, such as flow visualization or multivariate data visualization---these are driven by typical visualization needs in science and engineering. The VizCorner also tries to bring in trends from the visualization research community that might have a more long-term impact on science and engineering. For example, advanced human-computer interaction techniques might eventually improve how engineers or scientists work with their visual-interactive tools.

The VizCorner provided overviews of trends in visualization and the impact on computing in science and engineering, with synopsis of the 2011 and 2016 IEEE VIS Conferences (wwww.ieeevis.org), the premier venue for visualization research~\cite{Scheidegger2011,Comba2017}. The 2016 IEEE VIS report identified some trends that are still relevant. Let us have a closer look at current trends in visualization research.

\section{Trends in visualization}
